% Patch for section 04 (results_blind_benchmarks.tex): add real-data pan-cancer
% anchor paragraph after Shekhar retina pilot and before macaque retina / rank
% correlation. All numbers from agent/computational_biology @ba2c2e0 panel_A_tight_v2 27 rows.
%
% Intended splice point: after `\paragraph{Shekhar 2016 retina, BC8\_9 bipolar cells (real-data pilot).}`
% and before `\paragraph{Macaque retina, OFFx bipolar cells.}`

\paragraph{Cheng 2021 pan-cancer myeloid, LAMP3\textsuperscript{+} mregDC (real-data scale-series and method-regime anchor).} \emph{Fig.~3 ED + Panel A appendix 9-bar.} Committed at \texttt{agent/computational\_biology@ba2c2e0} with full evidence pack at \texttt{workspace/results/panel\_A\_tight\_v2.csv} (27 rows) and \texttt{workspace/results/locality\_grid.csv} (9 points). On Cheng 2021 PAAD ($n=2{,}853$, $\pi=1.07\%$), Cheng 2021 5-cancer pool ($n=20{,}341$, $\pi=1.50\%$), and Cheng 2021 6-cancer pool including KIDNEY ($n=49{,}271$, $\pi=1.15\%$), with LAMP3\textsuperscript{+} mregDC labels masked: all three whole-atlas operating points sit below the Theorem~1 multi-channel-composite floor ($n\pi^2\Delta^2 \in [0.2, 4.2]$ vs floor $9$), yet REAL fires at pan-cancer scale via sub-compartment $k$NN locality (§\Cref{rem:S1-effective-floor-locality}). Detection-strength $-\log_{10}(p_{\mathrm{bh}})$ under Harmony rises monotonically $0 \to 1.89 \to 2.39$ as $n$ grows 2.8k $\to$ 20k $\to$ 49k on the same held-out rare-type, validating Theorem~1's Le Cam local-sigmoid derivative. Under Scanorama the same LAMP3\textsuperscript{+} mregDC fires at all three scales ($p_{\mathrm{bh}} \in \{0.005, 0.008, 0.004\}$); under scVI also at all three scales ($p_{\mathrm{bh}} \in \{0.003, 0.004, 0.003\}$).

\paragraph{Three-regime empirical taxonomy.} The 9-point method $\times$ scale grid and the companion $\kappa$ per-motif cross-batch redistribution measurement (Supp.~Note~S\ref{supp:kappa}) reveal three integrator regimes on LAMP3\textsuperscript{+} mregDC, replicated on Zheng~2021 MM $\times$ CXCL13\textsuperscript{+} Tex at $n=12{,}274$ (orthogonal rare-type and orthogonal cancer):
\begin{itemize}\itemsep0pt
    \item \textbf{Regime A --- scale-mediated}: Harmony exhibits low $\kappa_{\text{median}} \approx 2.1$--$2.4$ and scale-monotonic detection (Theorem~1 scaling confirmed).
    \item \textbf{Regime B --- integrator-intrinsic mild modulation}: Scanorama exhibits high $\kappa_{\text{median}} \approx 8.3$--$9.1$ and fires at all scales regardless of $n$.
    \item \textbf{Regime C --- integrator-intrinsic scale-flat}: scVI exhibits medium $\kappa_{\text{median}} \approx 4.6$--$5.1$, fires at all scales, and its specificity is itself regime-dependent: at small-cohort curated scales (Sade-Feldman $n=16{,}291$, canonical TPEX) it preserves the motif at $p_{\mathrm{bh}}=1.0$ (mild regularisation); at pan-cancer Cheng6 scale its non-rare flag-rate rises to $0.647$ (scaling-induced feature-space dilution).
\end{itemize}
The three-regime stratification is reported as a \emph{descriptive empirical tool}, not a predictive theoretical taxonomy. A formally pre-specified cross-method unifying $\ell$-weighted regression prediction (Math Supp.~§\ref{rem:S1-effective-floor-locality}) was disconfirmed on the 9-point grid (workspace/results/locality\_regression\_v1\_negative.md); the scale-induced scaling and method-intrinsic aggressiveness mechanisms are therefore reported as distinct rather than unified. For graph-only integrators (BBKNN), the $\kappa$ framework requires a pre-specified point-embedding adapter; a retrospective prospective-test on BBKNN disconfirmed a diffmap-adapter $\kappa$ (phase-1) but confirmed an \texttt{X\_pca}-passthrough $\kappa$ (phase-2), establishing that adapter choice is part of the integrator specification for $\kappa$. For state-regime rare types distributed across motifs (Sade-Feldman canonical TPEX), motif-centroid $\kappa$ does not generalize; this is a documented scope limit, not a falsification of the core three-regime observation on cluster-regime rare types.

%% END PATCH 04 — splice before `\paragraph{Macaque retina, OFFx bipolar cells.}`
