\section{Introduction}

Single-cell RNA sequencing has grown over a decade from dozens of cells to atlases containing millions. The operation that makes this growth useful is \emph{batch integration}: the alignment of data across laboratories, sequencing platforms, dissociation protocols, and collection time points so that downstream analysis reflects biology rather than technical variation. Every integrator that the community has adopted --- Harmony, Seurat, scVI, scANVI, Scanorama, fastMNN, BBKNN, LIGER, scGen, scDML, scDREAMER, SCIDRL, scCRAFT, STACAS, CellANOVA --- occupies a point on a single tradeoff: the more aggressively batch variation is removed, the greater the risk that genuine biological variation is removed with it. This tradeoff is called \emph{overcorrection}, and it is the central tension in the integration literature \citep{scDML2023,scDREAMER2023,SCIDRL2022,scCRAFT2025,STACAS2024,CellANOVA2024,RBET2025,scIB-E2025}.

Rare cell populations are the first casualty of overcorrection. Their transcriptomic signal is weak relative to the scale of typical batch effects; their cells occupy a small volume in the embedding; and when integration aligns batches, they can be displaced into adjacent abundant populations and cease to exist as distinct biological entities. Multiple high-impact studies have documented this phenomenon on specific previously-discovered rare populations: pancreatic epsilon cells, pulmonary ionocytes, Schwann cells, activated stellate cells, OFFx bipolar cells in macaque retina, TCF7$^+$ stem-like CD8, LAMP3$^+$ mature dendritic cells, FOLR2$^+$ tumor-associated macrophages. A 2026 \emph{Nature Computational Science} review summarised that despite a decade of effort the field still oscillates between overcorrection and undercorrection. And yet the problem is not that overcorrection goes undetected --- it is that it is detected \emph{only} for populations the analyst already knew about.

The asymmetry has a second face: for five years the field has attempted to solve this problem and has not.  The failure is not for lack of attention.  It is structural.  Detecting loss of an unannotated rare subpopulation requires an evaluation framework; building such a framework requires a definition of success, which in turn requires ground truth for what was supposed to be preserved; and acquiring ground truth for populations that are by construction unannotated is precisely the step the framework was meant to replace.  No ground truth begets no evaluation, which begets no failure signal, which begets no motivation to develop a protective algorithm.  The loop is closed.  It is not, however, unbreakable: normal batch correction produces convergence (same-type cells merged across batches, cluster still present) whereas erasure of a rare subpopulation produces dispersion (cluster members scattered into unrelated large clusters), and these two patterns are structurally different in the data itself.  The signal exists; what was missing was a framework to extract it without external labels.

\input{handoffs/philo_framing_measured.tex}

\input{handoffs/philo_intro_para.tex}

\input{handoffs/immuno_intro_para.tex}

\input{handoffs/tumor_bio_intro_para.tex}

\input{handoffs/philo_onco_intro_para.tex}

Here we close the gap.  We introduce the \emph{Rare-subpopulation Erasure under Alignment, Label-free} (\textsc{REAL}) framework, which defines a rare subpopulation as a reproducible, multi-batch, high-density region in gene-expression space, and scores any integrator by how well it preserves such regions via a per-seed survival score, the LRS.  REAL does not consult annotation at any stage, which lets it detect loss of populations that were never named.  We derive \textsc{RareShield}, a protective integration procedure that anchors REAL-witnessed motifs in the latent space of any differentiable integrator without requiring labels.  We validate the pair on (i) label-masked positive controls where known rare populations are hidden from the pipeline, (ii) synthetic masked-unknown controls where rare populations are injected and then hidden, (iii) three immunology-anchored experiments (MAIT preservation across tissue batches, eTreg dilution in tumor atlases, pDC loss in severe COVID), (iv) a scale stress test on the Kang 2024 pan-cancer immune atlas with 4.9~million cells, and (v) cross-atlas replication of the flagged motifs on independent tumor atlases.  We close by arguing that the label-based benchmarks of the last decade should be extended with a label-free axis (which we call scIB-U), and that a fraction of the downstream conclusions built on existing integrated atlases should be re-examined under this axis.
