\section{Methods}

\subsection{Data and preprocessing}

Pancreatic dataset: Baron, Muraro, Segerstolpe, Wang pooled.  Epsilon cells removed from annotation before any step of REAL or RareShield is run on this dataset.  Airway dataset: Plasschaert / HLCA core with ionocyte labels removed.  Retinal dataset: macaque retina with OFFx labels removed.  Pan-cancer immune atlas: Kang et al. 2024 at full scale (4.9M cells).  Cross-atlas replication: HCA Lung Cancer and Gut Cell Atlas tumor arm.  Preprocessing: CPM-log normalisation; HVG selection at 4000 genes per dataset (consensus across batches).  Dimensionality reduction: batch-free PCA at 50 components; density estimation in PCA space.

\subsection{REAL hyperparameters (frozen)}

$k=15$ (per-batch kNN); $K=300$ (per-batch neighbourhood smoothing); $\tau=2.0$ (rarity anchor threshold); $c_{\min}=5$, $c_{\max}=0.005\cdot n_b$ (motif size window); $\mu=2$ (main text witness depth, $\mu=1$ in ED sensitivity); $\sigma_{\rm thresh}=0.85$ (cross-batch signature similarity).  Hyperparameters frozen on the positive-control split; not tuned on Kang.

\input{handoffs/ml_methods_integration_detection.tex}

\input{handoffs/math_methods_drop_in.tex}

\input{handoffs/math_hierarchical_real.tex}

\input{handoffs/immuno_methods_oracle.tex}

\input{handoffs/ds_methods_dataSci.tex}

\subsection{Benchmarks against 12 integrators}

Harmony, Seurat RPCA, Seurat CCA, scVI, scANVI, scVI + RareShield, scANVI + RareShield, Harmony-OT + RareShield, Scanorama, fastMNN, BBKNN, LIGER, scGen, scDML, scDREAMER, SCIDRL, scCRAFT.  Runs at default and at aggressive settings (Harmony $\theta\in\{1,2,4,6\}$; Seurat npcs $\in \{30,50,75\}$).  scIB metrics (ARI, NMI, ASW, graph connectivity, iLISI, cLISI, kBET) computed per the scIB reference implementation.  LRS is computed by our reference implementation at \texttt{agent/data\_science:workspace/lrs\_framework/}; integration harness at \texttt{agent/computational\_biology:workspace/code/}.  Runtime and memory profiled by \texttt{nvidia-smi} sampling.

\subsection{Immunology-anchored experiments (Exp-I1/I2/I3)}

As specified in \texttt{agent/immunology:workspace/immunology\_domain\_brief.md} \S5 and \texttt{rare\_immune\_subsets\_catalog.md} \S5.  Exp-I1: HCA Immune Cell Atlas + gut atlas; hold out non-blood tissue MAIT; measure LRS recovery.  Exp-I2: pan-cancer T cell atlas + Kang; dilution series of tumor eTreg.  Exp-I3 (main-text Fig.~6 panel): COMBAT COVID atlas; pDC prevalence collapse in severe COVID.

\subsection{Reproducibility}

Every run in the paper uses fixed random seeds; all seed values, hyperparameter sheets, and integrator versions are in Supp Table S3.  The frozen-hyperparameter sheet used for Tier A positive controls is committed at \texttt{agent/computational\_biology:workspace/code/configs/frozen\_tier\_a.yaml} with SHA recorded in every result file's \texttt{uns['method\_config']}.  Code, configs, and per-branch artefacts are released at \url{https://github.com/PoorOtterBob/eacn_example_001}.

\subsection{Statistics}

All confidence intervals are bootstrap percentile CIs over $\mathcal{W}$ motifs ($B=1000$) unless otherwise noted.  Integrator comparisons use Wilcoxon signed-rank tests on paired-motif LRS values.  Kang motif multiple-testing correction by Benjamini--Hochberg at $q=0.05$; conservative confounding-robust $p$-values use $\max(p_{\rm N1}, p_{\rm N3})$ across the batch-permutation and rotation nulls.
