% Methods §5 — Data-science contribution. Hand-off for Biological Science
% to splice into biological_science/workspace/manuscript/sections/07_methods.tex.
%
% Author: Data Science Agent (agent-mozutvsy), agent/data_science@504b41f.
% Nomenclature: REAL (framework), LRS (score), motif (entity), RareShield
% (protection). See Bio-Sci canonical naming broadcast (2026-05-10).

\subsection{Scope: rare, novel, distinct (Philosophy)}
\label{sec:methods-scope}

Adopted verbatim from Philosophy (\path{philo_methods_scope_clause.tex}):
``rare'' is count-based (frequency below $\pi_0$); ``novel'' is epistemic
(not named in a specified annotation reference); ``distinct'' is structural
(locally density-separated under a reproducibility test). REAL targets
populations that are rare and distinct irrespective of novelty: its output
does not depend on whether a population has a name. Within ``previously
unannotated'' we distinguish three nested reference classes: C1 novel-to-
dataset, C2 novel-to-literature, C3 novel-in-principle. REAL claims C2
\emph{bounded to} C3 via the exchangeability Lemma~S1. A REAL flag is
thus a C2 measurement claim, not a C1 dataset-completion claim or a C3
discovery claim.  Batch-rare singletons (present in one batch only) are
outside REAL's remit --- indistinguishable from batch-specific artefacts
by any cross-batch test.

\subsection{Per-motif survival score (LRS)}
\label{sec:methods-lrs}

For an integration map $f:\mathbf{X}_{\text{raw}} \to \mathbf{X}_{\text{int}}$,
each witnessed motif $w \in W$ with cell set $C_w$ has a survival score

\begin{equation}
\mathrm{LRS}(w;f) \;=\; \mathrm{HM}\bigl(P_d,\,P_t,\,P_n,\,1 - B\bigr),
\qquad
\mathrm{LRS}(f) \;=\; \mathrm{mean}_{w \in W}\,\mathrm{LRS}(w;f),
\end{equation}
where the four channels are, at default parameters $k_{\mathrm{kNN}} = 15$,
$k_{\mathrm{rarity}} = 300$, $\tau = 2.0$, $\sigma_{\mathrm{thresh}} = 0.85$,
$m_{\mathrm{wit}} = 2$:

\begin{itemize}
  \item $P_d(w) = \bigl[\rho_{\mathrm{int}}(C_w) / \rho_{\mathrm{pre}}(C_w)\bigr]_{[0,1]}$,
        the density persistence (local mass ratio).
  \item $P_t(w)$, the persistent-homology lifetime ratio: longest
        $H_0$-bar in a witness-complex filtration of $C_w \cup 5\lvert C_w\rvert$
        nearest non-motif cells, post vs. pre, clipped to $[0,1]$.
  \item $P_n(w)$, neighborhood integrity: fraction of $x \in C_w$ whose
        post-integration $k$-NN meets pre-integration $k$-NN within $C_w$.
  \item $B(w)$, compositional bleed: fraction of post-integration $k$-NN of
        $C_w$ that belong to the nearest abundant population.
\end{itemize}

The harmonic mean ensures that any channel collapsing to zero drives $\mathrm{LRS}(w)$
to zero; this is the test for total erasure. Thresholds: a motif is
\emph{eliminated} if $\mathrm{LRS}(w;f) < 0.4$ and \emph{preserved} if
$\mathrm{LRS}(w;f) > 0.7$, calibrated on the pancreas / HLCA / retina
hold-out validation set.

\subsection{Witness-complex persistent-homology scaling (Criterion 3)}
\label{sec:methods-scale}

At Kang-atlas scale ($n = 4.9 \times 10^6$, $q = 32$), the full Vietoris--Rips
complex is infeasible. We use a witness complex with
$m = \max(500, 2\lvert C_w\rvert)$ landmarks drawn from $C_w \cup 5\lvert C_w \rvert$
nearest non-motif cells. Stability is controlled by the landmark-selection
variance term from Chazal \& de Silva, plus Cohen--Steiner stability
(Mathematics Thm~2). $k$-NN construction uses FAISS-GPU; connected-component
enumeration on the motif-restricted graph uses cugraph. Witness-complex
filtration uses ripser++ on the landmark subsample. Empirical wall-clock:
$\le 2$ GPU-hours per 4.9M-cell evaluation on $8 \times$ A100.

\subsection{Cross-batch witness recovery (motif identifiability)}
\label{sec:methods-witness}

Motifs are identified from pre-integration embeddings only. Per batch
$b \in \{1, \ldots, B\}$: a per-batch rarity anchor
$R_b(x) = \rho_b(x) / \widetilde{\rho}_b(x)$ is computed, where $\rho_b$ is
$k$-NN density and $\widetilde{\rho}_b$ is the median of $\rho_b$ over a
larger $K$-NN neighbourhood. Seed clusters $S_b$ are $k$-NN connected
components restricted to $\{R_b > \tau\}$, size-filtered to
$[5,\,0.005 n_b]$. Cross-batch witness uses MNN on seed centroids in
HVG-cosine space with threshold $\sigma_{\mathrm{thresh}} = 0.85$ and
$m_{\mathrm{wit}} \ge 2$ distinct batch partners for main-text claims
($m_{\mathrm{wit}} = 1$ in ED1 sensitivity). Mathematics Thm~3 gives the
sample-complexity identifiability guarantee.

\subsection{Audit channels (not in headline LRS)}
\label{sec:methods-audit}

Two supplementary per-motif channels are emitted alongside the four-term LRS:

\begin{itemize}
  \item $P_\tau$: CoLM admissibility under RareShield's low-rank bounded-%
        Lipschitz batch-effect class $\mathcal{F}$, from Machine Learning.
  \item $P_\Delta$: Procrustes displacement ratio (post vs. pre centroid shift
        in HVG space, normalised by dataset global displacement), from
        Computational Biology (RareScore channel).
\end{itemize}

A third audit channel $D$ — \emph{disease-state-stratified neighborhood
purity} — is emitted only on disease-stratified atlases (Fig.~4 / Kang 2024),
proposed by Tumor Biology to surface cancer-type-specific erasure patterns
that a globally averaged $P_n$ would hide. $D$ is an audit column per
manuscript-lead decision; it is not folded into LRS.

\subsection{Label-free by construction; oracle firewall}
\label{sec:methods-firewall}

The LRS pipeline is label-free at every step: seed enumeration, cross-batch
witness, survival scoring. Known-rare marker panels supplied by Immunology
and Tumor Biology serve \emph{only} to compute an out-of-band
\texttt{oracle\_rare\_score} per cell, used downstream of LRS as the
ex-post grading signal in the Hide-and-Measure protocol
(Philosophy~\cite{philo}). A hard isolation check in the
\texttt{lrs\_framework.scoring} module rejects any run in which the
oracle-gene list intersects the HVG-based feature set used by LRS or
by any integrator. This forbids the Philosophy-flagged trap~1
(label leakage by proxy).

\subsection{Publication-grade visualization pipeline}
\label{sec:methods-viz}

All main figures are generated from reproducible matplotlib scripts under
\path{workspace/figures/figN_*.py}. Style is pinned in
\texttt{lrs\_framework.viz\_style}: Helvetica 7\,pt body text, Okabe-Ito
categorical palette, viridis sequential, PDF\,/\,PS font-type 42
(TrueType-embedded), vector everywhere except UMAP scatter which is
rasterised at 600\,DPI with vector axes. Accessibility is verified with
WCAG-contrast calculation and deuteranopia / protanopia simulation.
Each figure archives its generating script and input data hashes, so
every panel is regeneratable from raw AnnData via a single CLI call.

\paragraph{Data flow.}
Pre-integration AnnData $\to$ \texttt{lrs\_framework.precompute\_seeds}
(populates \texttt{obs['seed\_id\_pre']}, \texttt{obs['rarity\_R']},
\texttt{uns['lrs\_seeds']}) $\to$ integrator $\to$ per-method post-%
integration AnnData (preserves row order) $\to$
\texttt{lrs\_framework.scoring} $\to$ parquet of per-motif $\mathrm{LRS}(w)$
and audit channels $\to$ plotting scripts.
