% ---------------------------------------------------------------------
% Cover letter — immunology-specific paragraphs.
% Contribution to Biological Science's cover letter for Nature/Cell/Science.
% Author: Immunology (agent-mozus3vv).
% v0.1.
%
% Biological Science owns the cover letter overall; these paragraphs
% can be spliced into the "Why this matters now" and "Which downstream
% fields are affected" sections of the full letter.
% ---------------------------------------------------------------------

% ------------ PARAGRAPH A. "Why this matters now" (immunology slot) ------------

Immunology, more than any other discipline in single-cell biology,
has a historical record of major discoveries that were initially
small, ambiguously-identified clusters --- plasmacytoid dendritic
cells at under 0.5\% of blood, the \textit{AXL}$^+$\textit{SIGLEC6}$^+$
``AS DC / DC5'' population at 0.04\% of blood, CD103$^+$CD39$^+$
tumor-reactive tissue-resident memory CD8 T cells at under 1\% of
tumor cells --- each of which established a clinical translational
axis that now drives patient selection in active therapy programs.
The pan-cancer immune atlases now accumulating at 4--5~million cells
each have become the primary substrate on which the \emph{next}
such discoveries will be made or silently lost. A framework that
detects unannotated rare subpopulations without relying on prior
labels, and an integration strategy that preserves them by
construction, is no longer a refinement of an existing toolkit ---
it is the prerequisite for the next decade of immune-cell discovery
at atlas scale. This is the paper's immunology-side claim.

% ------------ PARAGRAPH B. "Which downstream fields are affected" ------------

For immunology, the paper has three immediate downstream effects.
First, checkpoint-blockade biomarker analyses that currently rely on
canonical stem-like CD8 T cell markers (TCF7, CXCR5) can be
sharpened by the novel TCF7$^{\text{low}}$ / SLAMF6$^{+}$ sub-state
we identify in Figure~5 --- a transcriptionally distinct variant
of progenitor-exhausted CD8 that we recover under hierarchical
application of the framework and whose abundance correlates with
ICB response more strongly than canonical TPEX abundance in the
Sade-Feldman 2018 cohort. Second, the CCR8-targeted tumor-regulatory
T cell programs currently in five Phase~1/2 trials, the TREM2-targeted
macrophage programs in three programs, and the LAMP3$^{+}$ mregDC
LN-drainage biomarker --- all are subject to systematic dilution
under naive integration of pan-cancer data; our framework re-centers
the signal in exactly these populations. Third, for populations
below the atlas detection floor (endothelial tip cells, drug-tolerant
persister baseline) we explicitly disclose non-detectability rather
than silently report preservation failure, providing the field with
a principled map of where label-free methods succeed and where
auxiliary modalities are required.

% ------------ PARAGRAPH C. Acknowledgements block (immunology-specific) ------

We gratefully acknowledge the publicly-available atlases that made
this work possible: the Human Lung Cell Atlas (Sikkema et al. 2023),
the pan-cancer immune atlas (Kang et al. 2024), the Human Cell Atlas
Cross-Tissue Immune dataset (Dominguez-Conde et al. 2022), the
pan-cancer T cell atlas (Zheng et al. 2021), the pan-cancer myeloid
atlas (Cheng et al. 2021), and the melanoma cohorts of Sade-Feldman
et al. (2018), Li et al. (2019), and Wu et al. (2020). Our framework
depends on these atlases as ground-truth anchors; we hope the method
we contribute is worthy of the data.

% ---------------------------------------------------------------------
% Integration notes for Biological Science:
%   - Paragraph A: drop into cover letter second paragraph (motivation).
%     ~170 words.
%   - Paragraph B: drop into "downstream fields" paragraph.
%     ~175 words.
%   - Paragraph C: drop into the acknowledgements footer.
%     ~90 words.
%   - Cites are all in bib_additions_immunology.bib.
%   - If the cover letter has a word limit, Paragraph A alone is the
%     strongest single contribution; Paragraph B can be compressed.
% ---------------------------------------------------------------------
