% ---------------------------------------------------------------------
% Methods fragment — oracle-score verification protocol.
% \input-ready for sections/09_methods.tex under a subsection like
% "Oracle-score verification of rare-subpopulation preservation".
% Author: Immunology (agent-mozus3vv). v0.1.
% ---------------------------------------------------------------------

\subsection{Oracle-score verification of rare-subpopulation preservation}
\label{sec:methods-oracle-score}

Verification of rare-subpopulation preservation after integration must
be strictly out-of-band: the marker-gene panels that score a cell's
identity post-integration cannot be allowed to influence the integration
itself, or the positive control collapses into circularity. We enforce
this by a separation of responsibility: the integration pipeline sees
only batch labels and (for semi-supervised methods such as scANVI or
Seurat-RPCA) a non-rare subset of the published cell-type annotation;
it never receives the marker-gene panel or any derivative statistic.
Marker-gene scoring occurs only after the integrated embedding has
been written to disk.

\paragraph*{Marker panels.}
We compile canonical positive-core, positive-supportive, and exclusion
marker sets for every candidate rare subpopulation (Supplementary
Table~\ref{tab:S3-marker-panels}). Positive-core markers are genes
expected to be expressed in essentially every cell of the target
subpopulation (for example \emph{CLEC4C}, \emph{LILRA4}, \emph{IL3RA},
\emph{TCF4}, \emph{IRF7} for plasmacytoid dendritic cells).
Positive-supportive markers are expressed in a majority of cells but
are not required (for example \emph{GZMB} in resting pDC, which
separates pDC from naïve B cell contaminants).
Exclusion markers identify the dominant false-positive neighbour
population (for example \emph{MS4A1} for B cells; \emph{CD1C} for
cDC2; \emph{CD3D} / \emph{CD3E} for T cells). Panels are constructed
from the published atlas-annotation literature
\citep{villani2017single, dominguez2022crosstissue, cheng2021pancancer}
and validated against CyTOF and CITE-seq cohorts where available.

\paragraph*{Oracle score.}
For each cell $x$ and panel $p$ with positive-core set $P_c$,
positive-supportive $P_s$, and exclusion $P_e$, we compute
\begin{equation}
  s_p(x) \;=\; \frac{1}{|P_c|} \sum_{g \in P_c} \hat{z}_g(x)
  \;+\; \tfrac{1}{2}\, \frac{1}{|P_s|} \sum_{g \in P_s} \hat{z}_g(x)
  \;-\; 2 \cdot \max_{g \in P_e} \hat{z}_g(x)
  \label{eq:oracle-score}
\end{equation}
where $\hat{z}_g(x)$ is the per-cell z-score of gene $g$ on log-normalized
counts (Seurat-standard \verb|pp.normalize_total| + \verb|pp.log1p| + \verb|pp.scale|).
A cell is oracle-positive for panel $p$ if $s_p(x) \geq 1.0$ and $s_p(x)$
exceeds the 95$^\mathrm{th}$ percentile across all cells in the dataset.
Both conditions are required to admit a cell to the oracle-positive set;
the upper-percentile condition prevents false positives on small datasets
where the absolute score threshold alone can be met by a non-trivial
fraction of cells.

\paragraph*{Firewall between panels and integration.}
We implement the firewall as a file-system boundary. Integration inputs
reside at \verb|shared/data/| in AnnData~H5AD format; marker panels
reside at \verb|shared/panels/| as JSON; the scoring code at
\verb|shared/detections/oracle/| reads both and writes only to
\verb|shared/integrations/*/scores/|. Integration code has read access
to \verb|shared/data/| but not to \verb|shared/panels/|. Test: every
integration run is executed with \verb|shared/panels/| unreadable
(enforced via directory permissions in the containerised runtime) and
the runs must succeed without the panels, confirming no implicit read
dependency.

\paragraph*{Preservation metrics.}
Given oracle-positive cells $O_p$ for each panel $p$ and the integrated
embedding $\phi(x)$, we compute (a) the fraction of $O_p$ whose Leiden
cluster in the integrated embedding is dominated ($\geq 70\%$) by
oracle-positive cells, (b) the per-cell $k$-nearest-neighbour purity of
$O_p$ in $\phi$ with $k=15$, and (c) the adjusted Rand index between
oracle-positive assignment and the integrated-embedding Leiden
partition restricted to $O_p \cup \mathrm{neighbour}(O_p)$. The three
metrics are reported jointly, with (a) as the primary measure for the
in-text claim and (b), (c) as supporting diagnostics in the supplement.

\paragraph*{Permutation null.}
For each panel $p$ we generate a size-matched random cell set $O_p'$
by sampling cells uniformly from the dataset, and compute the same
three metrics. Repeated $B = 1000$ times, these metrics define a
null distribution; an observed metric on $O_p$ that does not exceed
the 95$^\mathrm{th}$ percentile of the null is not reported as a
preservation success regardless of its absolute magnitude.

\paragraph*{Hierarchical (telescoping) application for below-whole-atlas populations.}
For rare populations whose whole-atlas prevalence lies below the
Le~Cam detection floor of \Cref{sec:minimax-detectability}, we apply
REAL-based oracle verification hierarchically. A population's effective
prevalence in a restricted sub-compartment (for example, pulmonary
ionocytes within the airway-epithelial subset of HLCA, or
CXCR5$^+$TCF7$^+$ TPEX within the exhausted-CD8 subset of Kang~2024)
is higher by the compartment-restriction factor
$n_{\text{whole}}/n_{\text{sub}}$, typically $10$--$100\times$.
Mathematics Supplementary Proposition~S1.6 (\emph{sub-compartment
detectability amplification}) formalises the condition under which
the restricted analysis admits detection. Our protocol applies
oracle verification at three nested scales: (i) the whole atlas,
(ii) a lineage-restricted subset (for example, all CD8 T cells,
or all epithelial cells), and (iii) a sub-lineage subset where
applicable (for example, TPEX within exhausted CD8). The primary
preservation metric for a population is reported at the scale at
which the $\pi\cdot\Delta^2$ product first exceeds the
$1.5\times10^{-4}$ detection floor, with sub-compartment analysis
cited explicitly. This telescoping protocol rescues populations
that are rare in the whole atlas but concentrated in a specific
lineage; it is required for ionocyte detection in the full HLCA
and for TPEX-sub-state detection in pooled pan-cancer melanoma
datasets.

% ---------------------------------------------------------------------
% Cross-references needed:
%   tab:S3-marker-panels (Supp Note S3, immuno_S3_markers.tex)
%   sec:evidence-package (evidence card schema discussion)
%   sec:loss-mode-taxonomy (Math's three objects)
% Bib entries needed:
%   villani2017single, dominguez2022crosstissue, cheng2021pancancer
%   (all in immuno_bib.bib)
% ---------------------------------------------------------------------
