% Mathematics agent — Good-Turing / Chao / Valiant connection brief.
% Companion to Philosophy's Supp Note on unseen-species estimators.
% Author: Mathematics Agent (agent-mozusggu), 2026-05-11.
% Drop into biological_science/workspace/manuscript/supplement/ as a short paragraph
% in the Discussion or appended to Supp Note S1.

\subsection{REAL in the unseen-species tradition}
\label{sec:S1-unseen-species}

REAL's exchangeability bound (Lemma~\ref{lem:exchangeability}) inherits the epistemic structure of a long tradition in statistics for estimating what has not been observed:
\begin{itemize}[leftmargin=*,nosep]
  \item \textbf{Good (1953)}, \textit{The population frequencies of species and the estimation of population parameters}: the Good-Turing estimator for the probability mass of unseen species, derived under an exchangeable-sampling assumption on species identities. Bound-valued, not point-valued.
  \item \textbf{Chao (1984, 1989)}: nonparametric lower bounds on total species richness using singletons and doubletons. Bound-valued by construction.
  \item \textbf{Valiant and Valiant (2011, 2013, 2016)}: unseen-distribution estimation with minimax-optimal sample complexity $\Theta(k/\log k)$ for $k$-distribution support. Bound-valued and matched to information-theoretic lower bounds.
\end{itemize}
The common structure across these: \emph{what we can certify about the unseen is a bound derived from exchangeability of the seen with the unseen.} REAL's Lemma~\ref{lem:exchangeability} is the single-cell analogue: held-out-known-rare loss rate upper-bounds truly-unknown loss rate, modulo a profile-mismatch term $\eta_{\text{match}}$. The minimax lower bound (Theorem~\ref{thm:minimax}) gives the information-theoretic floor beneath which no label-free method can succeed, matching the Valiant tradition.

The philosophical consequence: REAL's honest claims are bound-valued (we can assert ``rare motifs of mass $\pi$ and separation $\Delta$ such that $n\pi^{2}\Delta^{2} \geq 9$ are detected with power $\geq 0.95$''), not point-valued (we cannot assert the exact number of unseen motifs in Kang 2024). This matches the 70-year statistical pedigree of unseen-species estimation and addresses the reviewer objection ``how can you claim to find what you do not know?'' \citep{good1953,chao1984,valiant2011}.
