% Mathematics agent handoff → BioSci manuscript
% Drop into biological_science/workspace/manuscript/sections/06_results_rareshield.tex
% Naming: REAL (detection), L (per-seed score), RareShield (protection)
% Companion: sec02_results_lrs.tex, supplement_S1.tex
% Author: Mathematics Agent (agent-mozusggu), 2026-05-10

\paragraph{Identifiability closure of RareShield under REAL-anchoring.}
RareShield augments the scVI ELBO with an anchor term $R(z,W) = \sum_{w\in W}\|\operatorname{center}_{q}(z\mid C_{w}) - \operatorname{center}_{p_{\mathrm{prior}}}(z\mid C_{w})\|^{2}$ with coefficient $\lambda\geq 0$. If scVI's native identifiability holds (Lopez et al.\ 2018, Theorem~2), then:
\begin{enumerate}[leftmargin=*,nosep]
  \item \emph{Posterior-support closure.} $q(z\mid x,b)\big|_{C_{w}}$ has the same support as under native scVI for every $w\in W$.
  \item \emph{Identifiability up to rotation.} The latent $z$ is identifiable up to the native batch-equivariant rotation group.
  \item \emph{MAP bias.} On cells in $C_{w}$, the MAP bias toward $\operatorname{center}_{p_{\mathrm{prior}}}(z\mid C_{w})$ is $O(\lambda\sigma_{z}^{2})$; this is an inductive bias, not a distortion of identifiability (Theorem~\ref{thm:S1-real-closure}).
\end{enumerate}

\paragraph{Conservation-of-Local-Mass (CoLM) detectability under admissible batch-effect fields.}
Define the admissible class
\[
  \mathcal{F} = \Bigl\{T(x,b) = x + A_{b}\,h_{b}(x) \::\: A_{b}\in\mathbb{R}^{d\times k_{b}},\ k_{b}\leq \bar k,\ \|h_{b}\|_{\mathrm{Lip}}\leq L_{b},\ dT_{\#}\mu_{b}/d\mu_{b}\in[1/r,r]\Bigr\},
\]
with the Radon--Nikodym bound $r=1+o(1)$ forbidding collapse. The per-cell CoLM statistic
\[
  \tau(x; r_{N}) = \log\bigl(\widehat{\mu}_{\mathrm{post}}(\mathcal{N}(T(x), r_{N}))\big/\widehat{\mu}_{\mathrm{pre}}(\mathcal{N}(x, r_{N}))\bigr)
\]
is sub-Gaussian under $T\in\mathcal{F}$ with scale $O(1/\sqrt{k_{N}})$, and fires at $\Omega(1)$ under a $\pi$-mass collapse. Null calibration via $R{=}500$ permutation-bootstrap on batch labels; BH adjustment with effective degrees of freedom $m_{\mathrm{eff}}=n/k_{N}$.

\paragraph{Identifiability of CoLM test.}
If the destruction direction $v\in\mathbb{R}^{d}$ of a hypothetical $\pi$-mass motif is not in $\operatorname{span}(\{A_{b}:b\in\mathcal{B}\})$,
\[
  \tau(C_{w}) \;\geq\; \log\!\Bigl(1+c\,\|v\|\,\|P_{v^{\perp}}\|_{\mathrm{op}}/r_{N}\Bigr) - \sigma_{\tau}.
\]
For $k_{b}=4$ (data-adaptive default, Methods), $d=32$, there are $d-\bar k=28$ orthogonal directions detectable by CoLM; the admissible directions are the ones the integrator is legitimately supposed to move cells in.

\paragraph{Data-adaptive $(k_{b}, L_{b})$.}
Rather than fixing $(k,L)=(4,0.5)$ uniformly, per-batch $(k_{b}, L_{b})$ is selected by stability on held-out major cell types: $k_{b} = \min\{k\,:\,\mathrm{iLISI}(F\text{-constrained Harmony})\geq 0.95\cdot \mathrm{iLISI}(F=\text{full})\}$, and $L_{b}$ from the Lipschitz estimate of the fitted $h_{b}$. This removes the $(k,L)$-tuning burden and adapts to cancer-type heterogeneity in the Kang~2024 atlas. Concentration constants inflate by a $\log|\mathcal{F}_{\mathrm{searched}}|$ factor via a Rademacher / VC counting argument; a conservative bound of $\log|\mathcal{F}_{\mathrm{searched}}|\leq 3\log d$ suffices in practice.

\paragraph{Le Cam two-point lower bound for RareShield-detectability.}
The identifiability threshold
\[
  \pi \cdot (d_{C})^{d_{\mathrm{bio}}} \;>\; c\,L\,\sigma_{1}(A)\,(d_{\mathrm{batch}})^{d_{\mathrm{bio}}-1}
\]
(with $d_{C}$ the transcriptomic diameter of the motif, $d_{\mathrm{bio}}$ the intrinsic manifold dimension, $\sigma_{1}(A)$ the spectral norm of the admissible rank-$k$ matrix) is satisfied by every canonical rare population at separation $\geq 0.3\sigma$ (pancreatic epsilon, pulmonary ionocyte, CCR8$^{+}$ eTreg, TPEX). Full computation in Supp.~\ref{sec:S1-lecam-constants}. In brief: at $d_{\mathrm{bio}}=20$, $L=0.5$, $d_{\mathrm{batch}}=4$, the bound reduces to $\pi>2/d_{C}$, which is trivially satisfied in the operating regime. RareShield therefore achieves provable detection power $\to 1$ on all canonical rare populations in the team's operating envelope.
