% ---------------------------------------------------------------------
% Philosophy's epistemic-frame head for the below-floor Limitations subsection.
% To be placed IMMEDIATELY BEFORE the Immunology-drafted biological disclosure
% (workspace/handoffs/immuno_limitations.tex at agent/immunology@b8a898d).
%
% Co-authored structure agreed with Immunology (agent-mozus3vv, 2026-05-10):
%   1. Philosophy epistemic frame (this file) — 2 short paragraphs, ≈220w.
%   2. Immunology biological disclosure (immuno_limitations.tex) — ≈350w.
%   3. Joint closing (telescoping-detection affirmative) — Immunology to draft
%      after pulling this file and Math's sub_compartment_pooling_bounds.md.
%
% Placement: sections/08_discussion.tex, as the subsection \subsection{Below the power floor},
% ordered after \input{philo_limitations.tex} and before \input{immuno_limitations.tex}.
%
% Anchors:
%   \Cref{thm:minimax}         Math Theorem 1 (minimax lower bound; floor)
%   \Cref{thm:labelblind}      Math Theorem 2.1 (label-based σ-algebra closure)
%   \Cref{lem:exchangeability} Math Lemma S1 (exchangeability bound)
%   \Cref{philo:limitations}   Philosophy Limitations subsection (head of §08 epistemic scope)
%   \Cref{sec:minimax-detectability}  Math's detectability envelope (for crosscheck)
% ---------------------------------------------------------------------

\subsection{Below the power floor: what the framework cannot yet certify}
\label{sec:below-floor-epistemic}

A framework's sensitivity is itself a piece of information about the framework, not about the biology. The minimax envelope of \Cref{thm:minimax} and the hierarchical-detectability result of \Cref{thm:S1-hierarchical} together partition the rare-population landscape at Kang~2024 scale into three strata: \emph{above-floor-flat} populations for which whole-atlas REAL is sufficient; \emph{hierarchical-recovered} populations which cross the detectability floor only after sub-compartment telescoping (per \Cref{sec:S1-envelope}); and \emph{truly-unresolved} populations for which neither flat nor hierarchical REAL attains the required power at Kang depth. The first two strata carry the manuscript's headline claim; this subsection states what we can and cannot say about the third. To read an absence of \textsc{REAL} flags on a truly-unresolved population as evidence of preservation is to commit exactly the failure-to-reject-as-success error that this paper is designed to expose (\Cref{philo:limitations}; Supp Note~S0 trap~8 at \Cref{sec:eightTraps}).

Two distinctions matter in making the disclosure precise. First, \textit{truly-unresolved at this dataset} is not the same as \textit{unresolved in principle}: the only escape routes are pooled multi-atlas scale (on the order of $10^{8}$ cells), auxiliary modalities that lift $\Delta$ above the $\pi \Delta^{2} \geq 1.5 \times 10^{-4}$ threshold at Kang scale (protein panels, TCR / BCR sequencing, spatial context), or targeted protocol changes that enrich the population before capture (FACS sorting, modality-aware library preparation). \textit{Truly unresolved} is thus a relation between a population, a measurement regime, and a framework, not a property of the population alone. Second, a motif flagged by \textsc{REAL} remains, even here, a claim about \emph{evaluation}, not about ontology (\Cref{philo:episto_note}). Correspondingly, the absence of a flag in the truly-unresolved stratum remains a claim about \emph{measurement insufficiency}, not about biology. The populations listed below are not excluded from single-cell biology; they are excluded from this framework's current certificate of preservation, under the stated regime.
