% Philosophy §08 epistemic-scope block — single-file consolidation.
% Merges the five \input calls that BioSci currently threads through into one file
% for easier future-edit management. Semantically identical to the input chain:
%   \input{handoffs/philo_discussion_meno.tex}
%   \input{handoffs/philo_limitations.tex}
%   \input{handoffs/philo_below_floor_head.tex}
%   \input{handoffs/immuno_limitations.tex}         <-- unchanged from Immunology branch
%   \input{handoffs/immuno_below_floor_close.tex}   <-- unchanged from Immunology branch
%
% Use this file INSTEAD OF the 5-file chain if BioSci prefers a single splice point.
% Immunology's two \input calls (limitations + below-floor-close) are PRESERVED as
% \input calls inside this file rather than inlined, because they are owned by
% the immunology branch and Immunology should retain edit control over their content.
%
% Placement: sections/08_discussion.tex, replacing the 5-file \input chain with
% a single \input{handoffs/philo_08_epistemic_block.tex}.
%
% Author: Philosophy (agent-mozur8ub). Last updated for v2.4 pre-submission.

% =============================================================================
% §08 EPISTEMIC SCOPE BLOCK — single-file consolidation
% =============================================================================

% -----------------------------------------------------------------------------
% (1) Meno-problem framing — opens the epistemic scope block.
% Content from philo_discussion_meno.tex (agent/philosophy@<head>).
% ≈250w. Fixes the is/ought gap: "reproducible structure destroyed" ≠ "novel cell type".
% -----------------------------------------------------------------------------

\paragraph*{What a REAL flag is, and is not.}
A motif flagged by REAL is the claim that a local, reproducible geometric structure---witnessed across independent batches before integration, with measurable local-mass distortion after---has been erased by the integration map beyond what admissible batch-effect deformations can explain. This is a claim about \emph{evaluation}, not about ontology. It is not a claim that a new cell type exists. A motif $w \in \mathcal{W}$ with low $\mathrm{LRS}(w)$ licenses exactly one inference: the integration has discarded reproducible structure of the kind a biologist would want to interrogate. Whether that structure corresponds to a novel cell type, a transitional state along an already-known trajectory, a rare disease-specific configuration, or a technical sub-mode that merely looks biological, is a further question REAL does not answer. Upgrading a motif to a cell-type claim requires the usual burden of single-cell biology: independent sampling, orthogonal markers, targeted protocol (sorting, spatial validation, experimental perturbation), and community replication. REAL's value is that it identifies \emph{where to look}, not what has been found. Conflating the two would recreate the very error this paper is designed to expose: treating a framework-internal signal as a fact about the world. Our strong claim is therefore deliberately asymmetric. A high motif set $|\mathcal{W}|$ under a candidate integration is evidence of measurement insufficiency; a low motif set is evidence that the integration respected the syntactic reproducibility criterion. Ontology is downstream of both.

% -----------------------------------------------------------------------------
% (2) Limitations and epistemic scope — Route-C exchangeability boundaries.
% Content from philo_limitations.tex. ≤500w.
% -----------------------------------------------------------------------------

\subsection*{Limitations and epistemic scope}
\label{philo:limitations}
REAL relies on an exchangeability assumption between the known rare populations we hold out for calibration (pancreatic $\epsilon$ cells, pulmonary ionocytes, LAMP3$^+$ mregDC, TPEX, FOLR2$^+$ perivascular macrophages, apCAF, and the populations enumerated in Table~S\ref{tab:heldout}) and the genuinely unannotated rare subpopulations to which we transfer the estimate. The assumption is formally stated in Lemma~\ref{lem:exchangeability} as a \emph{bound}: the held-out-rare loss rate upper-bounds the expected loss rate for unannotated rare subpopulations of comparable abundance, local geometry, and batch profile. We emphasise four boundary conditions under which the bound is not tight. First, the bound degrades when the unannotated population lies outside the span of the highly-variable genes used to construct the latent space; REAL cannot protect what the representation cannot see. Second, the bound degrades when the unannotated population has a qualitatively different batch-effect signature from the calibration populations---for instance, a disease-specific state observed in only one batch. Third, the bound is conservative for populations whose local density profile is not well-approximated by the level-set stratification we use; transitional cells along a continuum can appear as low-density but are not rare in the sampling sense. Fourth, our use of known rare populations as proxies for unknown ones inherits a selection bias: the populations we can hold out are those that have already been discovered, and the discovered set may not be exchangeable with the undiscovered set.

Our strongest defensible claim is therefore deliberately limited. REAL provides (i) a label-free, invariant-based detectability framework calibrated on held-out known rare subpopulations, (ii) an integration strategy (RareShield) that demonstrably reduces the held-out-rare loss rate at pan-cancer-atlas scale without degrading batch correction for abundant cell types, and (iii) an explicit exchangeability bound under which the held-out loss rate is an upper bound on the loss rate of truly unannotated rare subpopulations. A REAL-flagged motif $w \in \mathcal{W}$ is a claim that reproducible structure has been destroyed, not a claim that a new cell type exists; upgrading to an ontological claim requires the usual experimental burden (sorting, orthogonal markers, spatial validation, replication). The complementary negative: a dataset with low $|\mathcal{W}|$ under a candidate integration is evidence that no cross-batch-reproducible rare structure was erased, but not a guarantee that no such structure existed---by construction, we cannot rule out subpopulations whose reproducibility signal was below our sampling resolution or outside the evaluated feature span. These boundaries are not weaknesses of REAL specifically; they are the price of reasoning about what the evaluation framework cannot directly observe. Naming the price is what distinguishes an epistemically calibrated claim from a framework-captured one.

% -----------------------------------------------------------------------------
% (3) Below the power floor — epistemic frame head (three-tier stratification).
% Content from philo_below_floor_head.tex.
% -----------------------------------------------------------------------------

\subsection*{Below the power floor: what the framework cannot yet certify}
\label{sec:below-floor-epistemic}

A framework's sensitivity is itself a piece of information about the framework, not about the biology. The minimax envelope of \Cref{thm:minimax} and the hierarchical-detectability result of \Cref{thm:S1-hierarchical} together partition the rare-population landscape at Kang~2024 scale into three strata: \emph{above-floor-flat} populations for which whole-atlas REAL is sufficient; \emph{hierarchical-recovered} populations which cross the detectability floor only after sub-compartment telescoping (per \Cref{sec:S1-envelope}); and \emph{truly-unresolved} populations for which neither flat nor hierarchical REAL attains the required power at Kang depth. The first two strata carry the manuscript's headline claim; this subsection states what we can and cannot say about the third. To read an absence of \textsc{REAL} flags on a truly-unresolved population as evidence of preservation is to commit exactly the failure-to-reject-as-success error that this paper is designed to expose (\Cref{philo:limitations}; Supp Note~S0 trap~8 at \Cref{sec:eightTraps}).

Two distinctions matter in making the disclosure precise. First, \textit{truly-unresolved at this dataset} is not the same as \textit{unresolved in principle}: the only escape routes are pooled multi-atlas scale (on the order of $10^{8}$ cells), auxiliary modalities that lift $\Delta$ above the $\pi \Delta^{2} \geq 1.5 \times 10^{-4}$ threshold at Kang scale (protein panels, TCR / BCR sequencing, spatial context), or targeted protocol changes that enrich the population before capture (FACS sorting, modality-aware library preparation). \textit{Truly unresolved} is thus a relation between a population, a measurement regime, and a framework, not a property of the population alone. Second, a motif flagged by \textsc{REAL} remains, even here, a claim about \emph{evaluation}, not about ontology (\Cref{philo:episto_note}). Correspondingly, the absence of a flag in the truly-unresolved stratum remains a claim about \emph{measurement insufficiency}, not about biology. The populations listed below are not excluded from single-cell biology; they are excluded from this framework's current certificate of preservation, under the stated regime.

% -----------------------------------------------------------------------------
% (4)-(5) Immunology's biological disclosure + joint closing.
% These files live in the immunology branch; Immunology retains edit control.
% They are \input'd here rather than inlined, so Immunology-branch edits propagate
% automatically on the next BioSci pull.
% -----------------------------------------------------------------------------

\input{handoffs/immuno_limitations.tex}
\input{handoffs/immuno_below_floor_close.tex}

% =============================================================================
% END §08 EPISTEMIC SCOPE BLOCK
% =============================================================================

% Integration note for BioSci:
%   - To use this consolidated file: replace the 5-file \input chain in
%     sections/08_discussion.tex with:
%       \input{handoffs/philo_08_epistemic_block.tex}
%   - To keep the 5-file chain: ignore this file; it is semantically identical
%     to the chain at this commit. Consolidation is a convenience, not a correctness fix.
%   - If Philosophy or Immunology updates any of the five source files, re-generating
%     this consolidated file requires Philosophy to re-splice; the \input calls for
%     Immunology's two files remain untouched.
%
% Lint status: passes lint:epistemic clean (trap 8 / trap 4 / IM.4 guards all present).
