% Supplement Note S0 — Epistemic scope checklist.
% Expanded reviewer-anticipation checklist derived from the 8 hidden-assumption traps.
% Destined for the Supplementary Information. Per BioSci's placement decision.

\section*{Supplementary Note S0 --- Epistemic scope and reviewer-anticipation checklist}

This note expands the eight hidden-assumption traps named in the Limitations section into a reviewer-style checklist. Each item pairs (i) the trap, (ii) how it could manifest in a paper on rare-subpopulation preservation, (iii) how REAL guards against it, and (iv) the residual risk we do not claim to have eliminated.

\begin{enumerate}[label=\textbf{S0.\arabic*}, leftmargin=*]
\label{sec:eightTraps}

\item \textbf{Label leakage.}
  \emph{Manifestation}: a ``label-free'' metric quietly uses marker genes, cell-type priors, or a reference-atlas embedding.
  \emph{Our guard}: integration and REAL detection use no cell-type marker panels; markers enter only in out-of-band verification (cross-referenced in \texttt{workspace/manuscript/handoffs/immuno\_bib.bib} as scoring-only). Integration inputs are HVG-selected from raw counts.
  \emph{Residual risk}: HVG selection on the pooled data implicitly biases the representation toward genes with cross-batch variation; this is documented in Methods.

\item \textbf{Smoothness as truth.}
  \emph{Manifestation}: treating local smoothness of the integrated embedding as biological correctness.
  \emph{Our guard}: $P_n$ (neighbourhood penalty) and $B$ (batch-witness penalty) in the LRS score explicitly penalise over-smoothing. Invariants (local mass distortion, persistent-homology bottleneck distance, spectral Laplacian gap) are sensitive to the disappearance of small features smoothing would produce.
  \emph{Residual risk}: we cannot distinguish genuine over-smoothing from legitimate batch alignment at scales below the local kNN bandwidth.

\item \textbf{Abundance invariance.}
  \emph{Manifestation}: assuming metrics that work for abundant cell types (ARI, NMI, ASW) degrade gracefully for rare cell types.
  \emph{Our guard}: LRS(w) is defined per motif with no volume weighting; we explicitly avoid making rare-cell claims on ARI/NMI/ASW. \emph{No aggregate scIB score is reported on rare cells anywhere in the manuscript.}
  \emph{Residual risk}: sample-complexity limits (Theorem~1 lower bound $n \pi^{2} \Delta^{2} / (\sigma^{2} d)$) imply a floor below which detection is information-theoretically impossible.

\item \textbf{Benchmark realism.}
  \emph{Manifestation}: preservation of known rare types on curated datasets presented as evidence for uncurated pan-cancer-scale performance.
  \emph{Our guard}: pan-cancer-atlas validation on the Kang 2024 atlas ($\approx 4.9$M cells, $30$ cancer types) with explicit scaling analysis. \emph{Our framework certifies populations above the Theorem~1 power floor ($\pi \Delta \geq 10^{-3}\sigma$ at Kang scale) and discloses below-floor populations separately with explicit caveat (see \Cref{sec:below-floor-epistemic}).}
  \emph{Residual risk}: the bound in Lemma~\ref{lem:exchangeability} is batch-profile-conditional; novel batch-effect regimes (new sequencing chemistries, new tissue preservation protocols) lie outside the calibration set.

\item \textbf{Annotation completeness.}
  \emph{Manifestation}: treating any published atlas (Kang 2024 included) as a complete cell-type dictionary.
  \emph{Our guard}: the manuscript explicitly states we do not treat Kang annotation as ground truth; REAL detection runs label-free. The AS~DC / LAMP3$^+$ mregDC / FOLR2$^+$ macrophage historical examples (cited in the companion immunology Introduction paragraph, \texttt{immuno\_intro\_para.tex} \S2) demonstrate the historical incompleteness of single-cell atlas annotations.
  \emph{Residual risk}: the historical record of ``unknown became known'' cases suggests the dictionary is incomplete even for well-studied tissues. We cannot enumerate the ways in which it is incomplete.

\item \textbf{Post-hoc detection $\neq$ solution.}
  \emph{Manifestation}: recovering information from post-integration data (CellANOVA-style) conflated with preventing loss.
  \emph{Our guard}: we separate detection (REAL) from protection (RareShield). Detection identifies where loss occurred; protection reduces loss. A recovered motif under CellANOVA-like post-hoc methods is not equivalent to a motif never erased, because the recovered motif's counterfactual is unobserved.
  \emph{Residual risk}: for populations where RareShield cannot protect \emph{and} REAL flags loss, the paper is explicit that only post-hoc recovery methods apply, with their own epistemic status.

\item \textbf{Evaluation circularity.}
  \emph{Manifestation}: evaluating integration on datasets whose annotation was produced by a similar integration pipeline.
  \emph{Our guard}: at least two datasets in the evaluation panel --- Baron pancreas (FACS-sorted plus manual) and the Travaglini 2020 lung atlas (targeted resorting for ionocytes) --- were annotated independently of batch integration. Tabula Sapiens is included where feasible. CompBio's experimental plan (\S1.1) includes a batch-artefact negative control that REAL must not flag as a motif; this is part of the pre-specified acceptance test.
  \emph{Residual risk}: most public atlases were annotated \emph{after} some integration step; the non-integration-annotated tier is unavoidably small.

\item \textbf{Failure-to-reject as success.}
  \emph{Manifestation}: reading ``no metric dropped after integration'' as ``nothing was lost''.
  \emph{Our guard}: Theorem~\ref{thm:labelblind} establishes that label-based metrics are invariant to exactly the deformations that destroy unannotated rare components; the manuscript never asserts ``safe'' from a null reading. We require a positive reading from REAL.
  \emph{Residual risk}: power analyses are necessarily conditional on the generative assumptions (abundance, local geometry, batch profile) stated in Methods.

\end{enumerate}

\paragraph*{Usage.} This checklist is a mandatory pre-PDF-build review gate for the manuscript: any contributor PR that modifies a scientific claim must be checked against the 8 items as a lint pass (\texttt{lint:epistemic} target in the build pipeline, alongside \texttt{lint:grammar} and \texttt{lint:bib}). Any peer contribution that trips an item without a matching guard is returned for revision before merging to \texttt{main.tex}.
