% ============================================================
% Introduction tumor-biology paragraph for sections/01_introduction.tex
% Owner: Tumor Biology Agent (agent-mozurh1r)
% Target: ~350 words, LaTeX-ready, to splice after the "why batch integration matters" paragraph
% and before the detectability-gap statement. Framework name: REAL. Protection: RareShield.
% ============================================================

\paragraph{Why cancer data is the worst case.}
The consequences of silent rare-subpopulation erasure are not uniform across biology. Cancer single-cell data is, by construction, the domain in which the conditions that make a population integration-fragile are stacked most tightly. Three properties co-occur. First, the populations that carry the most therapeutic weight in oncology are structurally rare: drug-tolerant persister cells are fewer than one percent of tumor cells at diagnosis \citep{Rambow2018Cell,Shen2020Cell,Oren2021Nature,Marine2020NatRevCancer}; LAMP3$^+$ mature regulatory dendritic cells are under three percent of tumor dendritic cells \citep{Maier2020Nature}; CCR8$^+$TNFRSF9$^+$ tumor effector regulatory T cells, the target of at least five Phase 1/2 programmes, are typically below one percent of the tumor microenvironment \citep{Plitas2016Immunity,DeSimone2016Immunity,Wang2019Cell,VanDamme2021,Kidani2022PNAS}. Second, these populations sit transcriptomically adjacent to their abundant neighbors: TCF7$^+$ stem-like CD8 T cells emerge from the exhausted CD8 cluster \citep{Miller2019NatImmunol,Siddiqui2019Immunity}; TREM2$^+$ lipid-associated macrophages sit inside the general tumor-associated-macrophage distribution \citep{Molgora2020Cell,Katzenelenbogen2020Cell}; antigen-presenting cancer-associated fibroblasts (apCAF) are routinely mislabeled as immune cells because of HLA-DR expression \citep{Elyada2019CancerDiscov}. Third, cancer cohorts are by construction non-exchangeable across batches: batch in cancer single-cell atlases partially encodes cancer type, treatment history, and disease state. Any integration procedure that enforces isotropy across batches therefore removes biology in proportion to how non-exchangeable the cohorts are.

The compounding effect is direct: the populations whose preservation most matters for precision oncology are the populations whose pre-integration signal is smallest, whose distance to their abundant neighbor is shortest, and whose cross-batch presence is the most asymmetric. Every existing integration benchmark has missed this failure mode because it measures preservation against labels that an earlier annotator already knew, and the populations at issue are frequently populations that annotators had not yet named. Here we resolve the gap: REAL supplies a label-free detectability framework that ranks reproducible rare motifs against a null of non-reproducibility, RareShield supplies a protective integration augmentation that reduces motif loss without compromising major-cell-type integration, and we apply both to the largest pan-cancer immune atlas currently available \citep{Kang2024PanCancerImmune} to estimate how often cancer rare biology has been silently collapsed under standard pipelines.
